A Theory for Database Transactions with no Database

### 2019-02-23 2018 Transactions/articleTransactions.tex
renewed sections 2 and 3

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\begin{document}

@@

\begin{abstract}
	How to do transactions if data is scattered across the internet?
	Users who must satisfy data integrity constraints are stuck with relational databases if they need transactions.
	For most NoSQL databases lack transaction functionality.
	This contribution proposes a theory of database transactions solely based on data,
	with no reference to any database in which that data resides.
	This makes transactions suitable for a wider class of databases than just relational ones.
	The soundness of the theory has been established by means of category theory.
	To show that the theory works, the theory is demonstrated by a prototype developed in Ampersand\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	To demonstrate that the theory is about real transactions, this contribution defines semantics of conventional database transactions.
	To fulfill our promise, we also gives semantics of transactions on the semantic web.
\end{abstract}

@@

\section{Introduction}
\label{sct:Introduction}

\section{Consistency}
\label{sct:Consistency}
	Each user of a database issues a stream of events.
	The order (in time) of events is given by a relation\footnote{%
A relation $\declare{r}{A}{B}$ is a set of tuples $\pair{a}{b}$ with $a\in A$ and $b\in B$} $\declare{pred}{Event}{Event}$,
	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.
	From a user perspective this stream is a sequence, so every event except the first has exactly one predecessor.

>>>>

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\def\id#1{\text{\em #1\/}}
\def\Events{{\mathit E}}
\begin{document}

@@

\begin{abstract}
	How can we make transactions work in distributed information systems without a shared database?
	Transaction theory as we know it assumes a shared data space.
	Multiple processes are acting simultaneously, each making changes to that shared data space.
	The observation that transaction mechanisms in NoSQL databases are not common
	makes us feel that a solution to this problem is not so obvious.

	This contribution proposes a theory of database transactions without a shared data space.
	This makes transactions suitable for more databases than just relational ones.
	The soundness of the theory has been established by means of formal proof.
	To show that the theory works, the theory is demonstrated by a prototype developed in Ampersand\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	To demonstrate that the theory is about real transactions, this contribution defines semantics of conventional database transactions.
	To fulfill our promise, we also gives semantics of transactions on the semantic web.
\end{abstract}

@@

\section{Introduction}
\label{sct:Introduction}
	A distributed information system (DIS) serves its users by registering data, guarding the consistency, protecting the data,
	and providing that data accurately and meaningfully for its intended purpose.
	Users want predictable behaviour when they are simultaneously changing a shared state (e.g.\ flight reservation).
	For that purpose a transaction mechanism preserves system-wide invariants.
	Many transaction mechanisms are known that presume a shared database [refs].
	However, shared databases are not always an option in distributed cloud architectures.
	Globally distributed applications such as Netflix and LinkedIn cannot afford to rely on a shared database for obvious reasons of performance and reliability.
	Many cloud applications use distributed storage models instead.
	Examples are event stores (e.g.\ Kafka), triple stores (e.g.\ semantic web), graph databases (e.g. Neo4J) and others.
	These so-called NoSQL databases often lack functionality for transactions.
	For the purpose of doing transactions in a non-transactional situation, several architectural patterns exist.
	Some of these patterns simply circumvent the problem.
	Others will isolate transactionality into a smaller sub-problem.
	Still others introduce locking mechanisms to do the job.
	Yet, in the world of event-driven distributed systems, a dire need is felt for a theory that brings what transactions do for shared databases.

	This paper presents such a theory.
	It describes transactions without the assumption of a shared database.
	In this theory we express the characteristic properties of transactions: Atomicity, Consistency, Isolation, and Durability.
	{\em Atomicity} means that a sequence of primitive actions can be executed as though it were a single action.
	{\em Consistency} means that integrity rules are warranted by the database management system.
	{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.
	{\em Durability} means that the contents of a database cannot change unintendedly.
	To motivate this theory, let us formulate requirements to characterize a DIS:

\section{Events}
\label{sct:Events}
	We need the notion of {em event}\footnote{Cambridge Dictionary defines ``event'' as: anything that happens} to describe things that happen in a DIS.
	Let us study the requirements for events in formal terms.
	In this contribution we use variables $e$, $e_1$, and $e_2$ to represent events.
	Let $\mathbb E$ represent the set of all conceivable events.

	For the purpose of discussing history, we introduce a structure that represents a history at one moment in time.
	A {\em historic structure} is a tuple $\la \Events, \id{prec}\ra$ in which
	\begin{eqnarray}
		&&\Events\text{ is a finite set of events.}\\
		&&\id{prec}\ \subseteq\ \Events\times\Events\label{req:precInE}\\
		&&\id{prec}^*\text{ is antisymmetric.}\label{req:precAntisymmetric}
	\end{eqnarray}
	Let set $\Events$ represent all events that the DIS has registered up to some point in time.
	So for every event $e\in\Events$ we can say that it has occured (at some time in the past).
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has occured before event $e_2$,
	making $\id{prec}$ a relation on events in $\Events$.
	\marginpar{claim}
	To represent time faithfully, we need an order on time that is a partial order,
	i.e.\ an order which is reflexive, transitive, and antisymmetric.
	For this reason requirement~\ref{req:precAntisymmetric} requires $\id{prec}^*$
	(which is the reflexive and transitive closure of $\id{prec}$)
	to be antisymmetric.
	With these requirements in place,  $\la\Events,\id{prec}^*\ra$ is the partial order that we need to represent time.

	An event $e$ that precedes another event is called a {\em past} event,
	which we define as \ $\exists e_1: e\ \id{prec}\ e_1$.
	All other events in a historic structure are called {\em current} events.
	The special case in which $\id{prec}^*$ is a total order is called a {\em linear historic structure}.
	Note that any nonempty linear historic structure has one current event.
	Other historic structures may have more.

	As time passes by, new events occur.
	We describe each occurrence as a mapping, $\id{occur}_{P,e}$, from one historic structure to another:
\begin{equation}
	\id{occur}_{P,e}(\la \Events, \id{prec}\ra)\ =\ \la \Events\cup\{e\}, \id{prec}\cup\{(e_1,e) | e_1\in P\}\ra
\label{req:def occur}
\end{equation}
	The occurrence of $e$ adds one event to $\Events$.
	It also enlarges relation $\id{prec}$ with one tuple for every immediate predecessor of $e$.
	The immediate predecessors of $e$ are specified in set $P$, which we call the predecessor set.
	Note that this enlargement can make current events become past events.
	If $h'=\id{occur}_{P,e}(h)$, the events in $P$ become past events in $h'$ and event $e$ becomes current in $h'$.

	We need the predecessor set to cater for the distributed nature of a DIS.
	In a linear historic structure,
\marginpar{claim}
	$P$ is alway a singleton set.
	In that case, every event except the first one has precisely one preceding event.
	In general, however, events may have multiple preceding events and the predecessor set is not necessarily a singleton set.

	Note that the definition of $\id{occur}_{P,e}$ (\ref{req:def occur}) must respect requirement~\ref{req:precInE}.
	If $P$ would contain events that are not in $\Events$, these events appear in relation $\id{prec}$ and hence violate requirement~\ref{req:precInE}.
\marginpar{claims}
	So this requires that $P\subseteq\Events$.
	Similarly, event $e$ cannot be an element of $\Events$, because that would violate the antisymmetry of $\id{prec}^*$ (requirement~\ref{req:precAntisymmetric}).
	So in a specification style with pre- and postconditions%
	\footnote{Also known as ``Floyd-Hoare style''}
	we specify:
\[\begin{array}{ll}
	\text{pre:}&P\subseteq\Events\\
	&e\notin\Events\\
	&\id{Events}=\Events\\
	&\id{pr}=\id{prec}\\[1em]
	\multicolumn{2}{l}{\ \ \ \id{occur}_{P,e}\la \id{Events}, \id{pr}\ra}\\[1em]
	\text{post:}&\id{Events}=\Events\cup\{e\}\\
	&\id{pr}=\id{prec}\cup\{(e_1,e) | e_1\in P\}\\
\end{array}\]

	We define $\la \Events, \id{prec}\ra$ a {\em historic substructure} of $\la \Events', \id{prec}'\ra$
	if $\Events\subseteq\Events'$ and $\id{prec}\subseteq\id{prec}'$.
	We denote this as
\begin{equation}
	\la \Events, \id{prec}\ra\ \subseteq\ \la \Events', \id{prec}'\ra
\end{equation}
\marginpar{claim}
	Note that $h'=\id{occur}_{P,e}(h)$ implies that $h\subseteq h'$.

	Having defined historic structures and the way they grow, let us proceed to determine requirements for threads and states.
	In the sequel, we will use variables $h$ and $h'$ to denote history structures.

\section{States}
	This section defines the requirements for predictable behavior of distributed information systems (DIS).
	First we need to introduce the notion of state.
	Let $\mathbb S$ represent the set of all conceivable states,
	with one special state $\mathcal I$ that represents the initial state.
	Let $\id{state}:\Events\rightarrow{\mathbb S}$ be a function that maps every event in a DIS to a state.
	We pronounce $\id{state}(e)$ as ``the state immediately after event $e$ has occured'',
	or shorter: ``the state after event $e$''.
	A {\em current state} of a DIS is the state after a current event of that system
	and a {\em past state} is the state after a past event.

	Each time a new event occurs, a change of state takes place.
	To accommodate such changes,
	we use triples $(e_1,e,\Delta)$ instead of tuples $(e_1,e)$.
	$\Delta$ represents the change from the state after event $e_1$ to the state after event $e$.
	Besides an event $e$ and immediate predecessors $P$,
	it takes a function $\delta$ to represent the change of state between two events:
\begin{eqnarray}
	\id{occur}_{P,e,\delta}\la \Events, \id{precd}\ra\ =\ \la \Events\cup\{e\}, \id{precd}\cup\{(e_1,e,\delta(e_1,e)) | e_1\in P\}\ra
\end{eqnarray}
	Additionally we require an operator $\oplus$ to compute the next state.
	This provides the means to define a {\em stateful historic structure} as a tuple $\la \Events, \id{precd}\ra$ in which
	$\id{precd}$ is a set of triples such that
\begin{eqnarray}
	&&\la \Events, \{(e_1,e)| (e_1,e,\Delta)\in\id{precd}\}\ra\text{ is a historic structure.}\\
	&&(e_1,e,\Delta)\in\id{precd}\ \Rightarrow\ \id{state}(e)=\id{state}(e_1)\oplus\Delta\label{eqn:nextState}\\
	&&(\exists e_1:(e_1,e,\Delta)\in\id{precd})\ \vee\ \id{state}(e)=\mathcal I\label{eqn:totState}\\
	&&\begin{array}{@{}l}
		(e_1,e,\Delta_1)\in\id{precd}\ \wedge\ (e_2,e,\Delta_2)\in\id{precd}\\
		\ \ \Rightarrow\ \id{state}(e_1)\oplus\Delta_1\ =\ \id{state}(e_2)\oplus\Delta_2\label{eqn:uniState}
	\end{array}
\end{eqnarray}
	Requirement~\ref{eqn:nextState} says that the state of non-initial events can be computed using $\oplus$.
	Requirement~\ref{eqn:totState} says that the state of an initial event (i.e.\ events without predecessor) is the initial state.
	Together, requirements~\ref{eqn:nextState} and~\ref{eqn:totState} define the function $\id{state}$ recursively.
\marginpar{claim}
	Requirement~\ref{eqn:uniState} ensures that $\id{state}$ is defined unambiguously,
	meaning that every event in $\Events$ has precisely one state.
	Thus, states are defined predictably in a DIS.

	We define $\la \Events, \id{precd}\ra$ a {\em stateful historic substructure} of $\la \Events', \id{precd}'\ra$
	if $\Events\subseteq\Events'$ and $\id{precd}\subseteq\id{precd}'$.
	We denote this as
\begin{equation}
	\la \Events, \id{precd}\ra\ \subseteq\ \la \Events', \id{precd}'\ra
\end{equation}
\marginpar{claim}
	Note that $\id{occur}_{P,e,\delta}(h)=h'$ implies that $h\subseteq h'$.

	\marginpar{claim}
	Can we prove $\oplus$ to be associative, commutative, and idempotent? Why would we do that?

	\section{Threads}
	So how does a stateful historic structure represent a distributed information system?
\marginpar{claim}
	Part of the answer is that a substructure of a stateful historic structure is a stateful historic structure by itself.

	The answer is that we have made no assumptions about location,
	like for instance that there is a shared memory.
	Even if we assume that each state exists in one location,
	a stateful historic structure can still have multiple current states.
	Each of these states can exist in a separate location.

	The purpose of threads is to support deterministic computations.
	We consider any nonempty historic structure that is totally ordered to be a thread.
	A {\em thread} is a historic structure $\la \Events, \id{prec}\ra$ that satisfies
\begin{eqnarray}
	\forall e_1,e_2&:&e_1\in\Events\ \wedge\ e_2\in\Events\ \Rightarrow\ e_1\ \id{prec}^*\ e_2\vee e_2\ \id{prec}^*\ e_1\label{eqn:order is total}\\
	\exists e&:&e\in\Events\label{eqn:nonempty thread}
\end{eqnarray}
	The total order (\ref{eqn:order is total}) is required to get deterministic results from a computation in a thread.
	In order to talk about the first and the last event in a thread, we require at least one event (\ref{eqn:nonempty thread}) in a thread.
	We shall use variable $t$ to represent threads.
	The first event in a thread $t$ is the infimum of the events in $t$;
	the last event is the supremum.

	We need the notion of state%
	\footnote{Cambridge Dictionary meaning of ``state'':
	a condition or way of being that exists at a particular time}
	to describe things that change over time.
	We shall use variables $s$, $s_1$, and $s_2$ to denote states.
	Let $\id{state}(e)$ represent the state after event $e$ has occured.

\section{Distributed Information System}

	Now equation~\ref{eqn:eval step} is the recursion step and the initial state is
	the starting point of the recursive computation of the final state of $t$.
	We consider a distributed information system as a set of processes that share a history.
	The system is distributed because processes need not share all histories.
	We need the notion of process%
	\footnote{Cambridge Dictionary meaning of ``process'':
		a series of actions that you take in order to achieve a result}
	to describe how components of a DIS interact.
	Let $\mathbb P$ be the set of all processes.
	We have formulated the following requirements:
	\begin{itemize}
	\item Belong/Contain\\
		To represent the fact that an event {\em belongs to} a process or
		(equivalently) that a process {\em contains} certain events,
		let $e\in p$ represent the statement that event $e$ belongs to process $p$.
	\item Shared history\\
		To abstract away from location, we make no assumptions with respect to the place where a process lives.
		However, to partake in a distributed system, processes must be able to share history.
		For that purpose we define what it means to be part of a history:
		\begin{equation}
			p\ \id{partof}\ h\ \Leftrightarrow\ (\forall e: e\in p\Rightarrow e\in h)
		\end{equation}
		We define an order between processes that share a history.
		\begin{equation}
			p_1\ \preceq\ p_2\ \Leftrightarrow(\forall e: e\in p_1\Rightarrow e\in p_2)
		\end{equation}
		We can prove that
		\begin{equation}
			p_1\ \preceq\ p_2\ \Rightarrow\ (\exists h:p_1\ \id{partof}\ h\wedge p_2\ \id{partof}\ h)
		\end{equation}
	\item Total order\\
		We require a process to have predictable behaviour, for which we need the process to contain a series (sequence) of events.
		So we require that $(\{e|e\in p\}, \preceq)$ is a total order.
		Apart from being a partial order, $\preceq$ must satisfy (for every $e_1$, $e_2$, and $p$):
		\begin{equation}
			e_1\in p\ \wedge\ e_2\in p\ \Rightarrow\ e_1\preceq e_2\vee e_2\preceq e_1
		\end{equation}
	\end{itemize}

\section{States}
	\begin{itemize}
	\item Revert\\
		A change can be used to compute a previous state.
		For this purpose we require an operator $\ominus$ such that
	\begin{equation}
		s_1\oplus\Delta=s_2\ \Leftrightarrow\ s_1=s_2\ominus\Delta
	\end{equation}
		As a consequence, we have for every event $e$:
	\begin{equation}
		\id{state}(\id{past}(e))=\id{state}(\id{pres}(e))\ominus\id{delta}(e)
	\end{equation}
		With $\preceq$ being reflexive, antisymmetric, and transitive,
		it follows immediately that $\oplus$ is idempotent, commutative, and associative.
	\end{itemize}

\section{Atomicity}
\label{sct:Atomicity}
	{\em Atomicity} means that multiple primitive actions are executed as one.
	So let us first define the notion of atomic action.

\section{Consistency}
\label{sct:Consistency}
	{\em Consistency} means that integrity rules are warranted by the database management system.

	If $s$ is a state and $r$ is an integrity rule, then $s\ \id{sat}\ r$ means that $s$ satisfies rule $r$
	and $s\ \overline{\id{sat}}\ r$ means that $s$ does not satisfy rule $r$.

	Let us call the set of integrity rules $\mathcal R$.
	A state $s$ is called consistent if it satisfies every integrity rule $r\in\mathcal R$
\begin{equation}
	\id{consistent}(s)\ =\ \forall r\in\mathcal R: s\ \id{sat}\ r
\end{equation}
	The system is consistent if the initial state is consistent and it can only reach consistent states by means of atomic actions.
	So all states in the history graph are consistent with $\mathcal R$.

\section{Isolation}
\label{sct:Isolation}
{\em Isolation} means that actions can be done without any knowledge of what other users are doing simultaneously.

\section{Durability}
\label{sct:Durability}
{\em Durability} means that the contents of a database cannot change unintendedly.

%% Obsolete

\section{Consistency}
\label{sct:Consistency}
Each user of a database issues a stream of events.
	The order (in time) of events is given by a relation\footnote{%
A relation $\declare{r}{A}{B}$ is a set of tuples. Each tuple $\pair{a}{b}$ in that set satisfies $a\in A$ and $b\in B$}
	$\declare{pred}{Event}{Event}$,
	where $\pair{e1}{e2}\in\id{pred}$ means that event $e1$ precedes event $e2$.
	From a user perspective this stream is a sequence, so every event except the first has exactly one predecessor.
